---
title: "Abortion Attitudes"
author: "Mahjoube AmaniChakani"
date: "2025-04-16"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(mc.cores = parallel::detectCores())
# load packages
library(rethinking)
library(brms)
library(lme4)
library(rmarkdown)
library(dplyr)
library(tidyverse)
library(labelled)
library(parameters)
library(mice)
library(tidyr)
library(viridis)
```


The data is online, and can be loaded using:

```{r data}
data <- read.csv('D:/scond year/QUANT/Project/data 7th/F00011356-WVS_Cross-National_Wave_7_csv_v6_0/WVS_Cross-National_Wave_7_csv_v6_0.csv')
head(data$incomeWB)
dim(data)
summary(data[c('Q184','I_ABORTLIB','Q262','Q261', 'Q260', 'incomeWB', 'Q273', 'H_URBRURAL', 'GII', 'Q37', 'Q50', 'Q274' )])


```


```{r missing values}

ab_data <- data %>%
  select(
    abortion_justif = Q184,        # 1–10
    age = Q262,                    # continuous
    sex = Q260,                    # 1 = Male, 2 = Female
    marital_status = Q273,         # 1–6
    urban_rural = H_URBRURAL,      # 1 = Urban, 2 = Rural
    duty_children = Q37,           # 1 = Strongly agree ... 5 = Strongly disagree
    fin_satisfaction = Q50,        # 1 = Dissatisfied ... 10 = Satisfied
    country = incomeWB,            # factor: Low, lower-middle, upper-middle, high
    GII = GII                      # continuous: 0–1
  ) %>%
  mutate(
    # Handle missing values for abortion attitudes
    abortion_justif = ifelse(abortion_justif %in% c(-5, -4, -2, -1), NA, abortion_justif),

    # Continuous predictors
    age = ifelse(age <= 0, NA, age),
    GII = ifelse(GII <= 0 | GII < -999, NA, GII),

    # Monotonic ordinal predictors (numeric, not factor)
    duty_children = ifelse(duty_children %in% c(-5, -4, -2, -1), NA, as.integer(duty_children)),
    fin_satisfaction = ifelse(fin_satisfaction %in% c(-5, -4, -2, -1), NA, as.integer(fin_satisfaction)),

  
    sex = factor(sex, levels = c(1,2), labels = c("Male", "Female")),
    marital_status = factor(marital_status),
    urban_rural = factor(urban_rural, levels = c(1,2), labels = c("Urban", "Rural")),
    country = factor(country)

  )





dropMissing_data <- ab_data[complete.cases(ab_data),]

head(dropMissing_data)

sum(is.na(dropMissing_data$abortion_justif))


summary(dropMissing_data$abortion_justif)
table(dropMissing_data$abortion_justif, useNA = "ifany")


```
#distribution
```{r dist table}
install.packages("skimr")
library(skimr)
skim(dropMissing_data_c)
table(dropMissing_data_c$age)

dropMissing_data_c <- dropMissing_data_c %>%
  mutate(age_group = cut(age,
                         breaks = c(0, 29, 44, 59, 100),
                         labels = c("18–29", "30–44", "45–59", "60+")))
dropMissing_data_c %>%
  count(age_group) %>%
  mutate(percent = n / sum(n) * 100)


dropMissing_data_c %>%
  count(sex) %>%
  mutate(percent = n / sum(n) * 100)

dropMissing_data_c %>%
  count(marital_status) %>%
  mutate(percent = n / sum(n) * 100)

dropMissing_data_c %>%
  count(urban_rural) %>%
  mutate(percent = n / sum(n) * 100)

dropMissing_data_c %>%
  count(duty_children) %>%
  mutate(percent = n / sum(n) * 100)


dropMissing_data_c %>%
  count(fin_satisfaction) %>%
  mutate(percent = n / sum(n) * 100)


```


#First part: abortion at in 10 categories

```{r initial graph}
simplehist(dropMissing_data$abortion_justif, xlim=c(1,10), 
           xlab="Attitudes toward abortion (1=Never justifiable, 5=neutral, 10=Always justifiable )")
simplehist(ab_data$abortion_justif, xlim=c(1,10), 
           xlab="Attitudes toward abortion (1=Never justifiable, 5=neutral, 10=Always justifiable )")
pr_k <- table( dropMissing_data$abortion_justif ) / nrow(dropMissing_data)

 
cum_pr_k <- cumsum( pr_k )  

 
plot( 1:10 , cum_pr_k , type="b" , xlab="Attitudes toward abortion (1=Never justifiable, 5=neutral, 10=Always justifiable )" , ylab="cumulative proportion" , ylim=c(0,1) )
logit <- function(x) log(x/(1-x)) 


round( lco <- logit( cum_pr_k ) , 2 )

```

#brms app
```{r prior_brms_1}
model_abortion <- bf(
  abortion_justif ~ mo(duty_children) + mo(fin_satisfaction) + age + sex + marital_status + urban_rural + GII + country,
  family = cumulative()
)


get_prior(model_abortion, data = dropMissing_data)



prior_abortion <- c(
  prior(normal(0, 0.5), class = "b"),
  prior(normal(0, 0.5), class = "Intercept")
)




prior_abortion

```



```{r model brms}

fit_abortion <- brm(
  formula = model_abortion,
  data = dropMissing_data,
  prior = prior_abortion,
  chains = 2,
  cores = 2,
  seed = 123,
  refresh = 0,
  iter = 400,
  warmup = 200,
  thin = 2,            
  threads = threading(4, grainsize = 100), 
  algorithm = "meanfield"
)


summary(fit_abortion,prob=0.89)


```
```{r}
fit_abortion %>% parameters( exponentiate = T, probs=0.95)

```

```{r pred_2}
# To see what the model predicts, we can create a newdata frame.
# For example, let’s compare predictions for different values of duty_children and fin_satisfaction.
cases <- data.frame(
  duty_children = c(1, 3, 5),
  fin_satisfaction = c(1, 5, 10),
  age = mean(dropMissing_data$age),
  sex = 1,              # 1 = Male, 2 = Female
  marital_status = 1,   # 1 Married
  urban_rural = 1,      # 1 = Urban
  GII = mean(dropMissing_data$GII),
  country=1 #Low income
)

cases

# Get predicted probabilities for each response category.
pred_abortion <- predict(fit_abortion, newdata = cases)
pred_abortion


cbind(cases, pred_abortion)
```


These visualizations reflect predicted ordinal response probabilities using line and bar plots with uncertainty (where applicable), consistent with the original style and expectations for Lab 14.
We can get the same thing with 90% credible intervals:

```{r pred_hisp_men_viz2}
gii_range <- seq(min(dropMissing_data$GII, na.rm = TRUE), 
                 max(dropMissing_data$GII, na.rm = TRUE), 
                 length.out = 1000)

newdata <- data.frame(
  duty_children = median(dropMissing_data$duty_children),
  fin_satisfaction = median(dropMissing_data$fin_satisfaction),
  age = mean(dropMissing_data$age),
  sex = 1,
  marital_status = 1,
  urban_rural = 1,
  GII = gii_range,
  country = 1
)

pred_probs <- posterior_epred(fit_abortion, newdata = newdata)
means <- apply(pred_probs, c(2, 3), mean)
lbound <- apply(pred_probs, c(2, 3), quantile, probs = 0.11)
ubound <- apply(pred_probs, c(2, 3), quantile, probs = 0.89)

plot(NA, xlim=range(gii_range), ylim=c(0,1),
     main="Predicted Abortion Justifiability by GII",
     xlab="Gender Inequality Index", ylab="Predicted Probability")

lines(gii_range, means[,1], lty=1)
lines(gii_range, means[,2], lty=2)
lines(gii_range, means[,3], lty=3)
lines(gii_range, means[,4], lty=4)
lines(gii_range, means[,5], lty=5)


polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,1],rev(ubound[,1])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,2],rev(ubound[,2])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,3],rev(ubound[,3])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,4],rev(ubound[,4])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,5],rev(ubound[,5])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,6],rev(ubound[,6])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,7],rev(ubound[,7])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,8],rev(ubound[,8])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,9],rev(ubound[,9])),
  col='#00000033', border=NA
)

polygon(
  x = c(gii_range,rev(gii_range)),
  y = c(lbound[,10],rev(ubound[,10])),
  col='#00000033', border=NA
)  
legend("topleft", lty = 1:10, lwd = 2,
       legend = c("1 = Never Justifiable", "2", "3", "4", "5 = Neutral",
                  "6", "7", "8", "9", "10 = Always Justifiable"),
       cex = 0.5)

```
```{r second graph:predict_and_visualize_gii_line_with_ci}

gii_range <- seq(min(dropMissing_data$GII), max(dropMissing_data$GII), length.out = 100)
dark <- "#23384F"
palette_colors <- viridis(10)

newdata_gii <- data.frame(
  duty_children = median(dropMissing_data$duty_children),
  fin_satisfaction = median(dropMissing_data$fin_satisfaction),
  age = mean(dropMissing_data$age),
  sex = 1,
  marital_status = 1,
  urban_rural = 1,
  GII = gii_range,
  country = 1
)

pred_probs_gii <- posterior_epred(fit_abortion, newdata = newdata_gii)
means_gii <- apply(pred_probs_gii, c(2, 3), mean)
lower_gii <- apply(pred_probs_gii, c(2,3), quantile, probs=0.11)
upper_gii <- apply(pred_probs_gii, c(2,3), quantile, probs=0.89)

plot(NA, xlim=range(gii_range), ylim=c(0,1),
     main="Predicted Abortion Justifiability by GII",
     xlab="Gender Inequality Index", ylab="Predicted Probability")

for (k in 1:10) {
  lines(gii_range, means_gii[,k], col=palette_colors[k], lwd=2)
  polygon(c(gii_range, rev(gii_range)),
          c(lower_gii[,k], rev(upper_gii[,k])),
          col=adjustcolor(palette_colors[k], alpha.f=0.4), border=NA)
}

legend("topleft", lty=1, lwd=1, col=palette_colors, c("1 = Never Justifiable", "2", "3", "4", "5 = Neutral",
                  "6", "7", "8", "9", "10 = Always Justifiable"), cex=0.5)


```




#Second part: abortion in 3 categories

```{r recoding}
dropMissing_data_c <- dropMissing_data %>%
  mutate(abortion_justif_3cat = case_when(
    abortion_justif %in% 1:3 ~ 1,  # Low justifiability
    abortion_justif %in% 4:7 ~ 2,  # Neutral
    abortion_justif %in% 8:10 ~ 3  # High justifiability
  ))
table(dropMissing_data_c$abortion_justif_3cat)

```


```{r initial graph new categories}
simplehist(dropMissing_data_c$abortion_justif_3cat, xlim=c(1,3), 
           xlab="Attitudes toward abortion (1=Never justifiable, 5=neutral, 10=Always justifiable )")
simplehist(dropMissing_data_c$abortion_justif_3cat, xlim=c(1,3), 
           xlab="Attitudes toward abortion (1=Never justifiable, 5=neutral, 10=Always justifiable )")


```
# Fit a new ordinal model using the recoded outcome

```{r model cat, paged.print=TRUE}
model_abortion_3cat <- bf(
  abortion_justif_3cat ~ duty_children + fin_satisfaction + age + sex + marital_status + urban_rural + GII + country,
  family = cumulative()
)

# Use the same priors
fit_abortion_3cat <- brm(
  formula = model_abortion_3cat,
  data = dropMissing_data_c,
  prior = prior_abortion,
  chains = 4,
  cores = 4,
  seed = 123,
  refresh = 0
)

summary(fit_abortion_3cat)

```





# Posterir prediction plot
```{r cat}

gii_range <- seq(min(dropMissing_data$GII, na.rm = TRUE), 
                 max(dropMissing_data$GII, na.rm = TRUE), 
                 length.out = 100)


newdata_gii <- data.frame(
  duty_children    = median(dropMissing_data$duty_children),
  fin_satisfaction = median(dropMissing_data$fin_satisfaction),
  age              = mean(dropMissing_data$age),
  sex              = 1,
  marital_status   = 1,
  urban_rural      = 1,
  GII              = gii_range,
  country          = 1
)


pred_probs_gii_3cat <- posterior_epred(fit_abortion_3cat, newdata = newdata_gii)

means_gii_3cat  <- apply(pred_probs_gii_3cat, c(2, 3), mean)
lower_gii_3cat  <- apply(pred_probs_gii_3cat, c(2, 3), quantile, probs = 0.11)
upper_gii_3cat  <- apply(pred_probs_gii_3cat, c(2, 3), quantile, probs = 0.89)

cols <- c("darkred", "gray30", "darkgreen")

# Plot setup
plot(NA, xlim=range(gii_range), ylim=c(0,1),
     main="Predicted Abortion Justifiability by GII (3 Categories)",
     xlab="Gender Inequality Index", ylab="Predicted Probability")


lines(gii_range, means_gii_3cat[,1], col=cols[1], lwd=2)
polygon(c(gii_range, rev(gii_range)),
        c(lower_gii_3cat[,1], rev(upper_gii_3cat[,1])),
        col=adjustcolor(cols[1], alpha.f=0.25), border=NA)


lines(gii_range, means_gii_3cat[,2], col=cols[2], lwd=2, lty=2)
polygon(c(gii_range, rev(gii_range)),
        c(lower_gii_3cat[,2], rev(upper_gii_3cat[,2])),
        col=adjustcolor(cols[2], alpha.f=0.25), border=NA)


lines(gii_range, means_gii_3cat[,3], col=cols[3], lwd=2, lty=3)
polygon(c(gii_range, rev(gii_range)),
        c(lower_gii_3cat[,3], rev(upper_gii_3cat[,3])),
        col=adjustcolor(cols[3], alpha.f=0.25), border=NA)

legend("topleft", lwd=2, lty=1:3, col=cols,
       legend=c("Low Justifiability", "Neutral", "High Justifiability"),
       cex=0.8)

```
# Visualization of Posterior Predictions

# Updated Multilevel Model (Random Intercepts for Country)

```{r fit_multilevel_model}
# Updated model with random intercept for country
model_abortion_multilevel <- bf(
  abortion_justif ~ duty_children + fin_satisfaction + age + sex + marital_status + urban_rural + GII + (1 | country),
  family = cumulative()
)
prior_abortion_multilevel <- c(
  prior(normal(0, 2), class = "b"),
  prior(normal(0, 5), class = "Intercept"),
  prior(exponential(1), class = "sd", group = "country")
)

start_time <- Sys.time()


fit_abortion_multilevel <- brm(
  formula = model_abortion_multilevel,
  data = dropMissing_data,
  prior = prior_abortion_multilevel,
  chains = 2,
  cores = 2,
  seed = 123,
  iter = 800,
  warmup = 200,
  thin = 2,            
  threads = threading(4, grainsize = 100), 
  algorithm = "meanfield"
)

end_time <- Sys.time()
end_time - start_time

summary(fit_abortion_multilevel)
plot(fit_abortion_multilevel, ask = FALSE)  # Check trace plots
pp_check(fit_abortion_multilevel)  # Posterior predictive checks
```

### Posterior Prediction Plot for GII (with Random Intercepts)

```{r predict_multilevel_gii}

gii_range <- seq(
  from = min(dropMissing_data$GII, na.rm = TRUE), 
  to = max(dropMissing_data$GII, na.rm = TRUE), 
  length.out = 1000
)


newdata <- data.frame(
  duty_children    = median(dropMissing_data$duty_children, na.rm = TRUE),
  fin_satisfaction = median(dropMissing_data$fin_satisfaction, na.rm = TRUE),
  age              = mean(dropMissing_data$age, na.rm = TRUE),
  sex              = 1,
  marital_status   = 1,
  urban_rural      = 1,
  GII              = gii_range,
  country          = 1  # reference country or average effect
)


pred_probs <- posterior_epred(fit_abortion_multilevel, newdata = newdata)


means  <- apply(pred_probs, c(2, 3), mean)
lbound <- apply(pred_probs, c(2, 3), quantile, probs = 0.11)
ubound <- apply(pred_probs, c(2, 3), quantile, probs = 0.89)

palette_colors <- RColorBrewer::brewer.pal(10, "RdYlBu")

plot(
  NA, xlim = range(gii_range), ylim = c(0, 1),
  main = "Predicted Abortion Justifiability by GII (Multilevel Model)",
  xlab = "Gender Inequality Index", ylab = "Predicted Probability"
)


for (k in 1:10) {
  lines(gii_range, means[, k], col = palette_colors[k], lwd = 2)
  polygon(
    c(gii_range, rev(gii_range)),
    c(lbound[, k], rev(ubound[, k])),
    col = adjustcolor(palette_colors[k], alpha.f = 0.2),
    border = NA
  )
}

legend(
  "topright", legend = paste("Score", 1:10),
  col = palette_colors, lwd = 2, lty = 1, cex = 0.75,
  title = "Justifiability", bty = "n"
)

```





```{r 2}
# Extract country-level intercepts
posterior <- as_draws_df(fit_abortion_multilevel)
ranefs <- posterior[, grepl("r_country\\[", names(posterior))]

# Extract the names and reformat
country_names <- gsub("r_country\\[|,Intercept\\]", "", colnames(ranefs))
colnames(ranefs) <- country_names

# Map country codes to income groups (make sure this exists in your data)
country_income <- data.frame(
  country = unique(country_names),
  income_group = c("Low", "Low", "High", "High")  # edit to match your countries
)

# Split posterior draws into two lists
low_income_draws <- ranefs[, country_income$country[country_income$income_group == "Low"]]
high_income_draws <- ranefs[, country_income$country[country_income$income_group == "High"]]

# Average across countries within group for each draw
posterior_mu_low <- rowMeans(as.matrix(low_income_draws))
posterior_mu_high <- rowMeans(as.matrix(high_income_draws))

# Plot comparison
xlim <- range(c(posterior_mu_low, posterior_mu_high))
ylim <- c(0, max(density(posterior_mu_low)$y, density(posterior_mu_high)$y))

plot(
  density(posterior_mu_low), 
  xlab = "Country-level intercept (latent scale)",
  ylab = "Density",
  xlim = xlim,
  ylim = ylim,
  main = "Posterior Distribution by Income Group",
  lwd = 2,
  col = "steelblue"
)
lines(density(posterior_mu_high), lty = 2, lwd = 2, col = "firebrick")

legend(
  "topleft",
  legend = c("Low-income countries", "High-income countries"),
  lty = c(1, 2),
  lwd = 2,
  col = c("steelblue", "firebrick"),
  bty = "n"
)
```



#3rd model
```{r sloe random}
model_random_slope <- bf(
  abortion_justif ~ duty_children + fin_satisfaction + age + sex +
    marital_status + urban_rural + GII + (1 + GII | country),
  family = cumulative()
)

# Priors (you can keep them weakly informative)
prior_random_slope <- c(
  prior(normal(0, 2), class = "b"), 
  prior(normal(0, 5), class = "Intercept"),  
  prior(exponential(1), class = "sd", group = "country"),  
  prior(lkj(2), class = "cor")  
)

# Fit with faster settings
fit_random_slope <- brm(
  formula = model_random_slope,
  data = dropMissing_data,
  prior = prior_random_slope,
  chains = 2,
  cores = 2,
  seed = 123,
  iter = 800,
  warmup = 200,
  thin = 2
)
```
```{r}

summary(fit_random_slope)


fixef(fit_random_slope)





```